Genome-wide changes in expression during inflorescence architecture development

### 2015-10-27 ms.Rmd
Add homeobox figure

Two results are presented here that suggest differing landscapes of expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM. Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *A. thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that this phenotype may be correlated with reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (see **ref next section**), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

Overexpression of the G1-Like (G1L/ALOG) transcription factor *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff]. *MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. Of the G1L genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*. Although the increase in secondary branches linked to *GIL5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from axillary mersitem to spikelet meristem. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki]. In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**].

A comparative analysis of expression of MIKC MADS genes in undifferentiated and differentiated meristems was used to investigate the evolutionary relationships between subclades in eudicots and monocots, exploiting two recent LMD RNA-sequencing datasets in addition to the one described in this paper. The RM and SM samples from rice inflorescences were compared with sequencing reads generated from the inflorescence meristem (IM) and FM of *A. thaliana* and the sympodial inflorescence meristem (SIM) and FM of the domesticated tomato, *Solanum lycopersicum* [@Park:2012dg; @Mantegazza:2014di]. A family- and clade-level overview of potential divergences, such as neo- and sub-functionalisation, was produced by combining a neighbour-joining similarity tree of MIKC MADS proteins with the change of expression of genes encoding them between indeterminate and determinate meristem tissues (`r I(pasteLabel("Figure ", figCount, "f_madsTree"))`). [**Note about relative genome size (number of genes) and MADS family size between species**.]

As expected based on their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems. However, a pair of tomato paralogs, apparently homologous to *SEP1* and *SEP2*, do not follow this pattern, which could indicate a role earlier or later in development or an unrelated function [**check SOLYC05G012020.2**].

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but the test statistic for the whole MADS family was higher in the SM than in the RM. An example of this is the MADS gene *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn]. In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of other MADS genes in those samples.

The strong expression of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the RM and PBM and their presence in clusters `r getClusterName('small', 'dec')` and `r getClusterName('dec', 'small')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM, gradually decrease in expression and have weakest expression in the SM. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc]. Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf]. WFP also binds to the promoters of *LOG1*, *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].

A homolog of *LONELY GUY 1* (*LOG1*), *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM. In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**]. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go]. Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O. sativa* ssp. *indica* varieties [@Ashikari:2005eg]. Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. *LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples. Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation from RM to SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level. A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.

@@

@@

>>>>

Two results are presented here that suggest significant changes in expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM. Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *A. thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

Overexpression of the G1-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*) increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. Yeast two-hybrid experiments indicate that G1L5 activates transcription of two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, but activation of the third rice *STMADS11*-like gene, *MADS47*, was not reported [@Yoshida:2013ff]. *MADS22* and *MADS55* also decrease in expression between PBM and SM, and *MADS55* is in the same cluster as *G1L5*, but *MADS47* is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). The presence and co-expression of genes that are involved in inflorescence development along with other uncharacterized TF genes indicates that the clusters capture biologically meaningful groups of genes. Of the G1L genes, which are annotated with the Pfam domain PF04852, only three were detected in the dataset, with two of them present in clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *G1L5* respectively), and the third, *G1L1*, not present in the clustering results but having a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *g1l5* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *G1L5* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed G1L genes, *G1L1*, *G1L2* and *G1L5*, with more than one being affected by the RNAi construct used against *G1L5*. Although the increase in secondary branches linked to *G1L5* suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *g1l5* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *G1L5*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem that remains to be reported.

@@

The patterns of differential expression of TF genes presented in `r I(pasteLabel("Figure ", figCount, "f_gsea"))` are not intended to represent each member of each gene family. For example,  the test statistic for the whole MADS family was higher in the SM than in the RM, but there are several MADS genes that have lower expression in the SM than in the RM, such as *PANICLE PHYTOMER2* (*PAP2*/*MADS34*), which was strongly expressed at all stages with higher expression in the RM and PBM. *PAP2* acts with another *AGAMOUS-LIKE2*-like (*AGL2*-like/*SEPELLATA*) MADS gene, *LEAFY HULL STERILE 1* (*LHS1*/*MADS1*), to specify floral organs [@Gao:2010iz], but *pap2* mutants also have altered primary and secondary branch number, suggesting an earlier role in spikelet meristem specification [@Gao:2010iz; @Kobayashi:2010hn]. In contrast, *LHS1*, which is involved in floral meristem determination and floral organ development [@Jeon:2000tj; @Agrawal:2005ho], was detected only in the SM samples and is thus representative the overall enrichment of MADS genes in the SM.

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity, and members of the *AGL2*-like clade interact with other MADS genes to specify floral organ identity, such as *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), which combine with *APETALA2* (*AP2*) to specify petals and another MADS gene, *AGAMOUS* (*AG*), to specify stamen [recently reviewed by @Prunet:2013ki]. In rice, functions homologous to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [**based on? check ref**], whilst stamen and carpels are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [**check, add ref**]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used LMD and RNA sequencing to study inflorescence development [@Park:2012dg; @Mantegazza:2014di]. Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *AGL6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [**ref Huang 2012**]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putatative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM, suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc]. Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [**ref Lu 2013**], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf]. WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [**ref Lu 2013**].

A homolog of *LOG1*, *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM. In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by **ref Han et al 2014**]. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM, small panicles with small SAMs and branching defects [@Kurakawa:2007go]. Similarly, reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*), which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O. sativa* ssp. *indica* varieties [@Ashikari:2005eg]. Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role. The LABA1 protein present in *O. rufipogon*, which has an identical primary sequence to the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. *LABA1* was detected in the ePBM/SBM and SM samples but not the RM or PBM samples. Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level. A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.

@@

## Orphan text (don't know where to put, consider deleting)

@@

```{r f_hb, fig.height = 8.661}
print(f_hb)
```

**`r I(pasteLabel("Figure ", figCount, "f_hb"))`**. Patterns of expression of homeobox genes in the meristem samples. The heatscale represents scaled, VST-transformed read counts for detected homeobox genes, and the coloured bar next to the *y*-axis shows the homeobox clade for each gene [**ref Jain et al. 2008**]. There is a set of homeobox genes that peak in expression in the RM, and another set that have their lowest expression in the RM and highest expression in PBM and SBM samples. The former group contains *OSH1* (**function?**), whilst the latter contains five of the eight expressed HD-ZIP IV homeobox genes.
